---
title: "Time Series Forecasting of Atmospheric CO2 using Prophet"
subtitle: "MTH6139 Time Series Coursework 1"
author: "Branden D'Souza"
date: "Spring term 2025"
output:
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
    highlight: tango
---

```{r, echo=FALSE}
# This code will display the QMUL logo at the top right of the page
# Do not change this code
htmltools::img(
  src = knitr::image_uri("images/QMlogo.png"),
  alt = "logo",
  style = "position:absolute; top:0; right:0; padding:10px; width:20%;"
)
```
# 1 Introduction

Time series analysis studies data that changes over time.
It is widely used in finance, economics and environmental science.

In my project I analysed atmospheric carbon dioxide (CO2)
levels using the built-in R dataset called `co2`.

My goal is to explore patterns in the data such as trends
and seasonal fluctuations and  use Meta's Prophet model
to forecast future CO2 levels.

## 1.1 About the Dataset

The dataset `co2` contains monthly measurements of atmospheric
carbon dioxide concentration in parts per million (ppm).

The measurements were taken at the Mauna Loa Observatory in Hawaii.

The dataset is commonly used in time series analysis because it
shows both a long-term increasing trend and clear seasonal patterns.

```{r}
start(co2)
end(co2)
frequency(co2)
```

The data is recorded monthly making it suitable for
time series forecasting as there is also lots of data making it make reliable.

## 1.2 Setup

```{r setup, message=FALSE, warning=FALSE}
library(prophet)
library(ggplot2)
library(zoo)
```

## 1.3 Visualisation

```{r}
plot(co2,
     main = "Atmospheric CO2 Concentration",
     ylab = "CO2 (ppm)",
     xlab = "Year")
```

The plot shows that atmospheric CO2 concentration increases over time and there is a repeating wave pattern across the series.Suggesting that the data contains both a long-term trend
and yearly seasonality.



## 1.4 Trend Analysis

The graph shows a clear upward trend in atmospheric CO2 levels.
The overall level rises steadily from around 316 ppm at the start
of the sample to over 365 ppm near the end.

Indicating that atmospheric CO2 concentration increased
consistently over the period observed.

A simple linear regression model is fitted
using time as the explanatory variable to support this.

```{r}
time_index <- 1:length(co2)

trend_model <- lm(as.numeric(co2) ~ time_index)

summary(trend_model)
```

The regression output shows a positive coefficient for time,
confirming that CO2 concentration increases over time.

This supports the conclusion that the series has a strong
positive long-term trend.


## 1.5 Seasonality Analysis

In addition to the long-term trend, the data also shows
a repeating seasonal pattern. The graph displays regular fluctuations that repeat each year. These appear as a wave-like pattern throughout the time series. This seasonal variation may be related to natural processes, such as plant growth cycles and seasonal changes in carbon absorption. To visualise the seasonal pattern more clearly, the series is decomposed into trend, seasonal and random components.

```{r}
co2_decomp <- decompose(co2)

plot(co2_decomp)
```

The decomposition separates the time series into three parts.

- Trend: the long-term increase in CO2 levels
- Seasonal: the repeating yearly fluctuations
- Random: irregular variation not explained by the model

The seasonal component confirms that the data contains
a strong yearly cycle with the same fluctuations throughout the decades.

## 1.6 Prophet Forecasting

Prophet requires the data to be stored in a dataframe with
two columns called `ds` and `y`.

The `ds` column contains the dates and the `y` column contains
the observed values.

```{r}
co2_data <- data.frame(
  ds = as.Date(as.yearmon(time(co2))),
  y = as.numeric(co2)
)

head(co2_data)
```

The Prophet model is then fitted to the CO2 data.

```{r}
co2_prophet_model <- prophet(co2_data)
```

To generate forecasts future monthly dates are created.
The model is used to forecast 60 months ahead, which is 5 years.

```{r}
future_dates <- make_future_dataframe(
  co2_prophet_model,
  periods = 60,
  freq = "month"
)
```

Forecasts are then produced for the observed and future dates.

```{r}
co2_forecast <- predict(co2_prophet_model, future_dates)

head(co2_forecast)
```

## 1.7 Forecast Plot

```{r}
plot(co2_prophet_model, co2_forecast)
```

The forecast suggests that CO2 levels will continue to rise
in the future. Upward rising trend.

The shaded region represents uncertainty around the predictions for the future.

## 1.8 Results Discussion

```{r}
prophet_plot_components(co2_prophet_model, co2_forecast)
```

The components plot shows the long-term trend and the yearly
seasonal pattern captured by the Prophet model.

The trend component remains upward-sloping, with the uncertainty with the shading after 2000's, while the yearly
component confirms the regular seasonal fluctuations in the data.

## 1.9 Conclusion

This project analysed the atmospheric CO2 time series using R
and Meta's Prophet forecasting model.

The data showed two main features: a strong upward trend and
clear yearly seasonality.

A linear regression model supported the presence of a positive
trend, while decomposition confirmed the repeating seasonal pattern.

The Prophet forecast suggests that CO2 levels are likely to
continue increasing over the next five years.
